Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-18 Manuscript/sec2-review-of-graph-neural-network.tex
Update section 2.

Different GNNs adopt different kinds of graph neurons and have different definitions of the three functions.
$\phi$ and $\Sigma$ are the \emph{edge calculation} functions.
They are conducted over every edge in $\mathcal{G}$.
$\gamma$ is the \emph{vertex calculation} function.
It is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.
For ChebNet, we report its GNN layer in the tables \footnote{A layer of ChebNet consists of $K$ GNN sub-layers and a summation layer, i.e., $\boldsymbol{H}^{l+1} = \sum_{k=1}^K{\boldsymbol{Z}^{(k)} \textcolor{blue}{\boldsymbol{W}^{(k)}}}$ with GNN layers $\boldsymbol{Z}^{(1)}=\boldsymbol{H}$, $\boldsymbol{Z}^{(2)}=\hat{\boldsymbol{L}}\boldsymbol{H}$, and $\boldsymbol{Z}^{(k)}=2\hat{\boldsymbol{L}}\boldsymbol{Z}^{(k-1)} - \boldsymbol{Z}^{(k-2)}$. $\boldsymbol{H}^l$ is the matrix of output hidden feature vectors of the layer $l-1$. In the table, we report the GNN sub-layer of ChebNet that calculates $\boldsymbol{Z}^{(k)}$}.

@@

						 & \text{Multi-head concatenation}: \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})                                                                                                                                             \\
						 & \text{Multi-head average}: \boldsymbol{m}_{j, i}^l = \frac{1}{K} \sum_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})

@@

			$O(K * d_{in} * d_{out})$                                                                                                   \\

@@

			$O(max(d_a, d_m, d_v) * K * d_{in})$                                                                                        \\

@@

	\caption{Typical graph neural networks and their edge calculation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer. Because two sum operator are interchangeable and sum operator and concatenation operator are interchangeable, we do it for GAT to explain more clearly in edge calculations functions.}

@@

			$\hat{\boldsymbol{h}}^{l,k}_{i} = \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_i^{l}$ (pre-edge), $\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$                                                         &

@@

			$O(max(K * d_v + d_{in}, 2 * d_{in} + d_m) * d_{out})$                                                                                                                                                             \\ \bottomrule

@@

        Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.7\columnwidth]{fig_GNN_complexity_quadrant.pdf}
	\caption{Complexity quadrants of typical GNNs. If two GNNs have the same complexity under the Big-O notation, we further compare their complexity according to how many parameters they need to train.}
	\label{fig:gnn_complexity_quadrant}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Neighbor sampling\label{fig:gnn_sampling_neigbhor_sampling}]{\includegraphics[height=4cm]{figs/illustration/layer_sampling.png}}
	%
	\subfloat[Graph sampling\label{fig:gnn_sampling_graph_sampling}]{\includegraphics[height=4cm]{figs/illustration/graph_sampling.png}}
	\caption{Training a GNN with sampling techniques. The faded graph neurons and their connections are inactivated.}
	\label{fig:gnn_sampling}
\end{figure}

>>>>

Different GNNs adopt different kinds of graph neurons and have different definitions of the three functions.
$\phi$ and $\Sigma$ are the \emph{edge calculation} functions.
They are conducted over every edge in $\mathcal{G}$.
$\gamma$ is the \emph{vertex calculation} function.
It is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.
For ChebNet, we report its GNN sub-layer in the tables. 
A ChebNet layer consists of $K$ GNN sub-layers and a summation layer:
$\boldsymbol{H}^{l+1} = \sum_{k=1}^K{\boldsymbol{Z}^{(k)} \textcolor{blue}{\boldsymbol{W}^{(k)}}}$ with the GNN sub-layers $\boldsymbol{Z}^{(1)}=\boldsymbol{H}^l$, $\boldsymbol{Z}^{(2)}=\hat{\boldsymbol{L}}\boldsymbol{H}^l$, and $\boldsymbol{Z}^{(k)}=2\hat{\boldsymbol{L}}\boldsymbol{Z}^{(k-1)} - \boldsymbol{Z}^{(k-2)}$, where $\boldsymbol{H}^l$ is the matrix of output hidden feature vectors of the layer $l-1$.
For GAT, a GAT layer consists of two sub-layers:
the first sub-layer calculates the summation $\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [\hat{\boldsymbol{h}}^{l,k}_i \parallel \hat{\boldsymbol{h}}^{l,k}_j] ))$, and the second sub-layer calculates $\alpha_{j, i}^k$ and $\boldsymbol{m}^l_{j,i}$.
The vertex calculation of a GAT layer is decomposed into two parts: $\hat{\boldsymbol{h}}^{l,k}_{i}$/$\boldsymbol{h}^{l+1}_i$ is calculated for every vertex before/after the message-passing, respectively.
The edge calculation of GAT uses $\hat{\boldsymbol{h}}^{l,k}_{i}$ during the message-passing.
For GaAN, a GaAN layer consists of four sub-layers: the first sub-layer calculates the summation $\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )$, and the other three layers calculate $\boldsymbol{m}^l_{j,i,1}$/$\boldsymbol{m}^l_{j,i,2}$/$\boldsymbol{m}^l_{j,i,3}$.

@@

						 & \text{Multi-head concatenation}: \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \hat{\boldsymbol{h}}^{l,k}_j)                                                                                                                                             \\
						 & \text{Multi-head average}: \boldsymbol{m}_{j, i}^l = \frac{1}{K} \sum_{k=1}^K \delta(\alpha_{j, i}^k \hat{\boldsymbol{h}}^{l,k}_j)

@@

			\makecell[r]{concat: $O(d_{in}*d_{out})$ \\
                 average: $O(K * d_{in} * d_{out})$  \\ Two sub-layers} 
             \\

@@

			\makecell[r]{$O(max(d_a, d_m, d_v) * K * d_{in})$ \\
                        Four sub-layers}\\

@@

	\caption{Typical graph neural networks and their edge calculation functions.
		Blue variables are model parameters to learn.
	}

@@

			\makecell[l]{$\hat{\boldsymbol{h}}^{l,k}_{i} = \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_i^{l}$ (pre edge calculation),\\ $\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$}                                                         &
			\makecell[r]{concat: $O(d_{in}*d_{out}/K)$\\ average: $O(d_{in}*d_{out})$
             \\ Two sub-layers}
			\\

@@

			\makecell[r]{$O(max(K * d_v + d_{in}, 2 * d_{in} + d_m) * d_{out})$\\
                         Four sub-layers}
            \\ \bottomrule

@@

		Blue variables are model parameters to learn.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.7\columnwidth]{figs/illustration/GNN_complexity_quadrant.pdf}
	\caption{Complexity quadrants of typical GNNs. If two GNNs have the same complexity under the Big-O notation, we further compare their complexity according to how many parameters they need to train.}
	\label{fig:gnn_complexity_quadrant}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Neighbor sampling\label{fig:gnn_sampling_neigbhor_sampling}]{\includegraphics[height=4cm]{figs/illustration/layer_sampling.pdf}}
	%
	\subfloat[Graph sampling\label{fig:gnn_sampling_graph_sampling}]{\includegraphics[height=4cm]{figs/illustration/graph_sampling.pdf}}
	\caption{Training a GNN with sampling techniques. The faded graph neurons and their connections are inactivated.}
	\label{fig:gnn_sampling}
\end{figure}
